Strong lensing selection effects

### 2023-01-30 paper/seleff.tex
final edits

%The luminosity and size distribution of strongly lensed sources can also be biased with respect to the general population of background galaxies.
%For instance, lensing magnification allows the detection of fainter objects with respect to the field, and, in case lenses are selected on the basis of the magnification of the source, galaxies that are very extended with respect to the Einstein radius are less likely to be classified as strongly lensed. This is because their magnification averaged over their size tends to be low \citep[see the discussion in section 5.3 of][]{O+A17}.

@@

The left-hand side of \Eref{eq:one} is directly accessible from strong lensing observations. If the main goal of a lensing survey is to characterise the properties of the strong lens population, then it can be accomplished by directly analysing this term. For many applications of strong lensing, however, the aim is to constrain the properties of the general galaxy or source population, $\prlens$ and $\prsource$, which are coupled in a nontrivial way via the lens selection probability $\psel$.
In order to obtain an unbiased estimate of either $\prlens$ or $\prsource$, then, it is necessary to invert \Eref{eq:one}.
In principle, this can be done with a Bayesian hierarchical formalism \citep{Son22}, but knowledge of the lens selection probability $\psel$ is required.
%For most of the existing strong lensing surveys, however, characterising $\psel$ is very challenging. %: the process of determining whether a system is included in a strong lens sample is typically a combination of several cuts, usually involving a nontrivial visual selection step. 
This factor can be written as the following product:
\begin{equation}\label{eq:two}
\psel(\psilens,\psisource|S) = \pdet(\psilens,\psisource)\pfind(\psilens,\psisource|S),%\footnote{\citet{Son22} implicitly assumed $\pfind\equiv1$, therefore $\pdet$ and $\psel$ can be used interchangeably in the context of that work},
\end{equation}
where $\pdet$ is the probability of detecting a strong lensing event, and $\pfind(\psilens,\psisource|S,\mathrm{det})$ is the probability of correctly classifying it as such\footnote{\citet{Son22} implicitly assumed $\pfind\equiv1$, therefore $\pdet$ and $\psel$ can be used interchangeably in the context of that work.}.
The detection probability $\pdet$ can be obtained via simulation. The main challenge is characterising $\pfind$: in most of the existing strong lensing surveys, the process of determining whether a system is included in a strong lens sample is typically a combination of several cuts, usually involving a nontrivial visual selection step.

For the above reasons, the problem of inverting \Eref{eq:one} is a difficult one to tackle exactly.
Few studies have attempted to explicitly account for strong lensing selection effects, usually by making ad-hoc simplifying assumptions \citep{Son++15,O+A17,Son++19}.
%The problem of the invertibility of \Eref{eq:one} can be simplified by adopting lens selection criteria that can be easily forward-modelled: for instance by building samples that are complete above a well-defined observational cut \citep{Son22}.
%The problem can be simplified by adopting lens selection criteria that can be easily forward-modelled: for instance by building samples that are complete above a well-defined observational cut \citep{Son22}.
Whether it is necessary to invert \Eref{eq:one}, however, depends on the severity of the strong lensing bias that needs to be corrected and on the accuracy requirements on the key quantities of interest.

In this paper we aim to quantify the strength of the strong lensing bias on a series of foreground galaxy and background source parameters. %, under realistic assumptions for the lens detection probability $\pdet$.
In particular, we aim to determine how strong lenses differ from the parent population of foreground galaxies and background sources in terms of a) the radial mass structure of the lenses (i.e. their stellar and dark matter mass density profiles); b) the ellipticity of the lenses; c) the size-magnitude distribution of the lensed sources.
The answer to this question depends on 1) how the lens detection probability $\pdet$ varies as a function of galaxy and source properties, 2) the efficiency of a survey at correctly classifying detected strong lenses (i.e. $\pfind$), and on 3) the shape of the galaxy and source parameters distribution $\prlens$ and $\prsource$. To understand this third point we can imagine the limiting case in which both $\prlens$ and $\prsource$ are Dirac delta functions (i.e. all lenses and sources are identical): in this limit, $\prsl$ simply reduces to the product $\prlens\prsource$ up to a scaling constant, corresponding to a case in which the lensing bias is none.

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisited the \citet{MVK09} study, expanding it to the extended source case. We simulated individual lenses and examined how the lens detection probability varies with lens and source properties.
In addition, we addressed points 2) and 3) as well: we simulated large populations of strong lenses using empirical models, we simulated the lens detection and finding phase, and quantified the lensing bias under various scenarios.
%We explore how the results change as a function of the scatter in mass parameters at fixed light, of the efficiency of finding lenses with small image separation, and of source type: we simulate both galaxy and quasar sources.
%We explore how the results change as a function of the scatter in mass parameters at fixed light, and as a function of the efficiency of a lens survey at discovering small image separation lenses.
We explored how the results change as a function of the efficiency of a lens survey at discovering small image separation lenses, and of the scatter in mass parameters at fixed light.

%In light of the results of our experiments, we discuss to what extent can studies of galaxy structure based on strong lenses be generalised to describe non-lens galaxies.
Finally, we addressed the question of how different are galaxy-galaxy lens samples from sets of galaxy-quasar lenses. This last point is relevant for time-delay cosmography studies, in which measurements of the time-delay between the multiple images of a strongly lensed quasar are used to constrain the expansion rate of the Universe. Galaxy-galaxy lenses can in principle be used to help break some of the model degeneracies affecting these measurements \citep{B+T21}, but any difference between the two lens classes can introduce biases, if not corrected for.
With this study we aim to quantify this bias.

The structure of this work is the following.
In \Sref{sect:basics} we introduce the basics of gravitational lensing.
In \Sref{sect:indlenses} we study individual lens systems. %, to look for trends in the lensing cross-section with various lens and source properties.
In \Sref{sect:lenspop} we describe our simulations of lens surveys.
In \Sref{sect:results} we show the results of our analysis on the simulated lens survey data.
We discuss the results in \Sref{sect:discuss} and draw conclusions in \Sref{sect:concl}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution $\kappa(\boldsymbol\theta)$ (also referred to as the convergence). 
This is the ratio between the surface mass density and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cr}}}.
\end{equation}
%which is the ratio between the projected surface mass density of the lens and the critical surface mass density for lensing.
The latter quantity is defined as
%The quantity $\Sigma_{\matrhrm{cs}}$ is the critical surface mass density for lensing, defined as
\begin{equation}
\Sigma_{\mathrm{cr}} = \dfrac{c^2D_\mathrm{s}}{4\pi G D_\mathrm{d} D_{\mathrm{ds}}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_\mathrm{d}$, $D_\mathrm{s}$, and $D_{\mathrm{ds}}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

In the special case of an axisymmetric lens we can simplify the notation by considering a single coordinate axis with origin at the centre of the lens. We label the components of the image position, source position and deflection angle along this axis as $\theta$, $\beta$ and $\alpha$, respectively.
The lens equation for an axisymmetric lens then becomes
\begin{equation}\label{eq:1dlenseq}
\beta = \theta - \alpha(\theta),
\end{equation}
and the expression for the deflection angle reduces to
%In the special case of an axisymmetric lens, the deflection angle, calculated along an arbitrary axis passing through the lens centre, reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:1dlenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form there have infinite magnification in the tangential direction.
%
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}
%Based on \Eref{eq:criteq},
%we generalise the definition of $\tein$ to the non-axisymmetric case as the circularised radius of the isodensity contour that encloses an average surface mass density equal to $\Sigma_{\mathrm{cr}}$.

Axisymmetric lenses of the kind considered in this work 
%Axisymmetric lenses with a density profile of the kind considered in section \ref{ssec:profile} 
typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
%\Fref{fig:scheme}, which shows the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, helps to visualise this property. 
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for a few lens models.
%for the same lens models considered in \Fref{fig:kappa}.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curve.

For small values of $\beta$ (for sources close to the lens centre), the number of images that are produced is three: the source is strongly lensed. For large values of $\beta$, instead, only one image is formed.
The value of $\beta$ where the transition occurs is known as the radial caustic, which is marked by the horizontal dotted line in \Fref{fig:scheme}. 
As can be seen from \Fref{fig:scheme}, a source at this location is mapped to the stationary point of the $\theta - \alpha(\theta)$ curve. 
%this value corresponds to the stationary point of the $\theta - \alpha(\theta)$ curve. 
That location on the image plane is known as the radial critical curve. 
Images that form there have a formally infinite magnification in the radial direction\footnote{The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the radial magnification. This can be understood by taking the derivative of \Eref{eq:1dlenseq} with respect to $\theta$.}.

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for five lenses with different density profiles. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
The lens models used in this simulations consist of a stellar component and a dark matter halo, as described in section \ref{ssec:profile}. Their density profile is plotted in \Fref{fig:kappa}.
\label{fig:scheme}
}
\end{figure}
%The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the magnification of an image in the radial direction. 
%This can be understood by noting that, according to \Eref{eq:1dlenseq}, the derivative of $\theta - \alpha(\theta)$ is equal to $d\beta/d\theta$, which quantifies the ratio between an infinitesimal displacement in the radial direction in the source plane and the corresponding displacement in the image plane.
%Stationary points of the $\theta - \alpha(\theta)$ curve, therefore, are points of infinite magnification in the radial direction. For a given axisymmetric lens, these points describe a circle known as the radial critical curve. 
%The corresponding curve in the source plane, the radius of which is given by the value of $\theta - \alpha(\theta)$ at the stationary point, is known as the radial caustic.
%As can be seen in \Fref{fig:scheme}, sources that lie within the radial caustic are strongly lensed into three images, while sources that lie outside of it produce only one image.

%Strong lenses also have a tangential critical curve, where the magnification along the tangential direction is formally infinite. For an axisymmetric lens, the tangential critical curve is a circle with radius equal to the Einstein radius, and the corresponding tangential caustic collapses to the $\beta=0$ point in the source plane.

@@

\Fref{fig:ellcaust} shows the source-plane caustics of elliptical lenses with different values of the axis ratio. We used the software {\sc Glafic} \citep{Ogu21} to obtain the caustic curves.
The outermost curves are radial caustics, while the inner ones are tangential ones.
%While in the axisymmetric case these are a circle and a point respectively (blue curves), in the elliptical case they are transformed into a 
The most striking difference with respect to the axisymmetric case (blue curves in \Fref{fig:ellcaust}) is the fact that the tangential caustic is transformed from a point into a diamond-like curve.
Sources located within the diamond produce five images (one of which is usually highly de-magnified). Sources that lie in the region enclosed between the two caustics are imaged three times. Sources outside the radial caustic are imaged only once, as in the axisymmetric case.
The fact that the number of images changes by two at a caustic crossing is a general feature of gravitational lenses with no singularities \citep{SEF92}.

@@

All of the examples that we considered were based on sources with a single surface brightness peak.
In general this is not necessarily the case, and the intrinsic number of surface brightness peaks of a strongly lensed source is not known a priori.
When dealing with a real lens candidate, applying our lens definition criterion requires showing that the observed number of surface brightness peaks can be reproduced with a lens model in which the source has a smaller number of peaks.
In samples of lenses defined via visual inspection this process is typically done implicitly, by identifying multiply imaged blobs that belong to the same source element.
%Applying our lens definition criterion to a real lens requires showing that the observed number of surface brightness peaks can be reproduced with a lens model in which the source has a smaller number of peaks.

To our knowledge, we are the first to propose a surface brightness peak-based definition of a strong lensing event.
A popular alternative definition of a strong lens is one based on magnification: only images that are magnified by more than a given threshold are considered as strongly lensed \citep[see e.g.][]{Hil++07}.
The problem with a definition of this kind is that magnification cannot be determined unambiguously from observations, unless the intrinsic properties of the lensed source are known from other means (e.g. if the source is a standard candle or a standard ruler).
Because of the mass-sheet degeneracy \citep{FGS85}, it is possible to vary the magnification of a lensed image while keeping its observed properties fixed: this could lead to the paradox of two identical-looking lenses that are classified differently on the basis of the underlying magnification.
Although we could still use a magnification-based definition for the sake of carrying out our experiments, it would then be difficult to apply our results to real data.
Our definition of a strong lensing event, instead, is robust with respect to the mass-sheet degeneracy.
%Because of the mass-sheet degeneracy

@@

%In this work we focus on massive elliptical galaxies as lenses. This is because this class of objects makes up the vast majority of known strong lenses.
%The type of lenses that we focus on in this work are massive elliptical galaxies: these make up the vast majority of known strong lenses.
%The lens galaxies that we focus on in this work are massive early-types: these make up the vast majority of known strong lenses.
In this work we focus on massive early-type galaxies as lenses, as these make up the vast majority of known strong lenses.
We describe their mass distribution with a model consisting of two concentric components, one describing the baryons and one for the dark matter.
We assume that the baryonic component consists entirely of stars, thereby neglecting gas, which is known to contribute very little to the mass of early-type galaxies in the inner regions that are probed by strong lensing. 
We then assume that the stars follow a S\'{e}rsic profile, with %with a fixed index of $n=4$ (also known as a de Vaucouleurs profile).
projected surface mass density given by
\begin{equation}
\Sigma(R) = \Sigma_0 \exp{\left\{-b\left(\frac{R}{\reff}\right)^{1/n}\right\}}.
\end{equation}
In the above equation,
\begin{equation}
\Sigma_0 = \frac{\mstar b^{2n}}{2\pi n \reff^2 \Gamma(2n)},
\end{equation}
$\mstar$ is the total mass, $\reff$ is the radius enclosing half of the total mass, $n$ is the S\'{e}rsic index, $\Gamma$ is the incomplete Gamma function, and $b$ is given by \citep{C+B99}
\begin{equation}
b(n) \approx 2n -\frac13 + \frac{4}{405n} + \frac{46}{25515n^2} + O(n^{-3}).
\end{equation}
Throughout this paper we indicate with $\teff$ the angular size of the half-light radius.

We fix the S\'{e}rsic index of the lenses to $n=4$, corresponding to a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is most relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. In section \ref{ssec:limitations} we discuss qualitatively what the implications of relaxing this assumption would be.
%This assumption is probably broken in real galaxies but, as we discuss in \Sref{ssec:limitations}, does not impact our investigation.

%Throughout this work we describe the mass distribution of the lens galaxies with a two-component model consisting of a S\'{e}rsic profile, describing the baryonic component of a galaxy, and a generalised Navarro, Frenk \& White (gNFW) profile, describing the dark matter halo.
%We assume that the two components are concentric.
%We also assume that the baryonic component consists entirely of stars.
%This is a good approximation for the majority of the known strong lenses, which are massive elliptical galaxies with little gas.
%Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. This assumption is probably broken in real galaxies but, as we discuss in \Sref{sect:discuss}, does not impact our investigation.

%The projected surface mass density of a S\'{e}rsic profile is given by
%\begin{equation}
%\Sigma(R) = \Sigma_0 \exp{\left\{-b\left(\frac{R}{\reff}\right)^{1/n}\right\}},
%\end{equation}
%with
%\begin{equation}
%\Sigma_0 = \frac{\mstar b^{2n}}{2\pi n \reff^2 \Gamma(2n)},
%\end{equation}
%where $\mstar$ is the total mass, $\reff$ is the radius enclosing half of the total mass, $n$ is the S\'{e}rsic index, $\Gamma$ is the incomplete Gamma function, and $b$ is given by \citep{C+B99}
%\begin{equation}
%b(n) \approx 2n -\frac13 + \frac{4}{405n} + \frac{46}{25515n^2} + O(n^{-3}).
%\end{equation}
%Throughout this paper we indicate with $\teff$ the angular size of the half-light radius.

@@

\Fref{fig:kappa} shows the dimensionless surface mass density profile of S\'{e}rsic + gNFW models with various values of the inner dark matter density slope $\gammadm$ and of the fraction of projected dark matter mass within the half-light radius, $\fdm$.
All of the profiles have %a S\'{e}rsic index $n=4$, 
a dark matter scale radius equal to ten times $\reff$, and are normalised in such a way that the Einstein radius is equal to the half-mass radius of the stellar component (these assumptions are dropped later).
Two main features emerge from \Fref{fig:kappa}. First, the baryons generally dominate the total density in the inner regions $(\theta < \tein)$, while the dark matter is the main component at large radii.
Second, models with different dark matter fractions and inner dark matter slopes can conspire to produce very similar total density profiles. 
This is the case, for example, of the ($\fdm=0.5,\gammadm=1.5$) and the ($\fdm=0.2,\gammadm=2.0$) models (green and blue lines in \Fref{fig:kappa}).

@@

Axisymmetric lenses with a density profile of the kind introduced in section \ref{ssec:profile} can produce either one or three images of a point source.
This can be seen in \Fref{fig:scheme}, which shows the lens equation for various values of the dark matter fraction and the inner dark matter slope.
All of the lenses shown in this figure have the same Einstein radius, which is equal in size to the half-mass radius of the stellar component.
The radius of the radial caustic, marked by the dotted lines in \Fref{fig:scheme}, is a strong function of the lens properties: it is largest in lenses with a smaller dark matter fraction or steeper dark matter slopes.
As a result, the source plane area that is subject to strong lensing is an even stronger function of these properties, since it scales with the square of the radial caustic radius.
In order to compute the lensing cross-section, however, we must take into account the magnification, because that determines whether multiple images can be detected or not.
%In order to compute the lensing cross-section, however, we must take into account the magnification of the multiple images: according to our definition of a strong lensing event, given in section \ref{ssec:lensdefpoint}, at least two images must be detected in order for a system to qualify as a strong lens.

>>>>

@@

The left-hand side of \Eref{eq:one} is directly accessible from strong lensing observations. If the main goal of a lensing survey is to characterise the properties of the strong lens population, then it can be accomplished by directly analysing this term. For many applications of strong lensing, however, the aim is to constrain the properties of the general galaxy or source population, $\prlens$ and $\prsource$, which are coupled in a nontrivial way via the lens selection probability $\psel$.
In order to obtain an unbiased estimate of either $\prlens$ or $\prsource$, then, it is necessary to invert \Eref{eq:one}.
In principle, this can be done with a Bayesian hierarchical formalism \citep{Son22}, but knowledge of the lens selection probability $\psel$ is required.
This factor can be written as the following product:
\begin{equation}\label{eq:two}
\psel(\psilens,\psisource|S) = \pdet(\psilens,\psisource)\pfind(\psilens,\psisource|S),
\end{equation}
where $\pdet$ is the probability of detecting a strong lensing event, and $\pfind(\psilens,\psisource|S,\mathrm{det})$ is the probability of correctly classifying it as such\footnote{\citet{Son22} implicitly assumed $\pfind\equiv1$, therefore $\pdet$ and $\psel$ can be used interchangeably in the context of that work.}.
The detection probability $\pdet$ can be obtained via simulation. The main challenge is characterising $\pfind$: in most of the existing strong lensing surveys, the process of determining whether a system is included in a strong lens sample is typically a combination of several cuts, usually involving a nontrivial visual selection step.

For the above reasons, the problem of inverting \Eref{eq:one} is a difficult one to tackle exactly.
Few studies have attempted to explicitly account for strong lensing selection effects, usually by making ad-hoc simplifying assumptions \citep{Son++15,O+A17,Son++19}.
Whether it is necessary to invert \Eref{eq:one}, however, depends on the severity of the strong lensing bias that needs to be corrected and on the accuracy requirements on the key quantities of interest.

In this paper we aim to quantify the strength of the strong lensing bias on a series of foreground galaxy and background source parameters. 
In particular, we aim to determine how strong lenses differ from the parent population of foreground galaxies and background sources in terms of a) the radial mass structure of the lenses (i.e. their stellar and dark matter mass density profiles); b) the ellipticity of the lenses; c) the size-magnitude distribution of the lensed sources.
The answer to this question depends on 1) how the lens detection probability $\pdet$ varies as a function of galaxy and source properties, 2) the efficiency of a survey at correctly classifying detected strong lenses (i.e. $\pfind$), and on 3) the shape of the galaxy and source parameters distribution $\prlens$ and $\prsource$. To understand this third point we can imagine the limiting case in which both $\prlens$ and $\prsource$ are Dirac delta functions (i.e. all lenses and sources are identical): in this limit, $\prsl$ simply reduces to the product $\prlens\prsource$ up to a scaling constant, corresponding to a case in which the lensing bias is none.

\citet{MVK09} carried out a thorough study of point 1): they quantified how the properties of a lens galaxy determine its probability of creating a lensing event with a point source.
In this work we revisited the \citet{MVK09} study, expanding it to the extended source case. We simulated individual lenses and examined how the lens detection probability varies with lens and source properties.
In addition, we addressed points 2) and 3) as well: we simulated large populations of strong lenses using empirical models, we simulated the lens detection and finding phase, and quantified the lensing bias under various scenarios.
We explored how the results change as a function of the efficiency of a lens survey at discovering small image separation lenses, and of the scatter in mass parameters at fixed light.

Finally, we addressed the question of how different are galaxy-galaxy lens samples from sets of galaxy-quasar lenses. This last point is relevant for time-delay cosmography studies, in which measurements of the time-delay between the multiple images of a strongly lensed quasar are used to constrain the expansion rate of the Universe. Galaxy-galaxy lenses can in principle be used to help break some of the model degeneracies affecting these measurements \citep{B+T21}, but any difference between the two lens classes can introduce biases, if not corrected for.
With this study we aim to quantify this bias.

The structure of this work is the following.
In \Sref{sect:basics} we introduce the basics of gravitational lensing.
In \Sref{sect:indlenses} we study individual lens systems. 
In \Sref{sect:lenspop} we describe our simulations of lens surveys.
In \Sref{sect:results} we show the results of our analysis on the simulated lens survey data.
We discuss the results in \Sref{sect:discuss} and draw conclusions in \Sref{sect:concl}.

@@

The lensing properties of an object with respect to a source depend solely on its dimensionless surface mass density distribution $\kappa(\boldsymbol\theta)$ (also referred to as the convergence). 
This is the ratio between the surface mass density and the critical surface mass density for lensing:
\begin{equation}
\kappa(\boldsymbol\theta) = \dfrac{\Sigma(\boldsymbol\theta)}{\Sigma_{\mathrm{cr}}}.
\end{equation}
The latter quantity is defined as
\begin{equation}
\Sigma_{\mathrm{cr}} = \dfrac{c^2D_\mathrm{s}}{4\pi G D_\mathrm{d} D_{\mathrm{ds}}},
\end{equation}
where $c$ is the speed of light, $G$ the gravitational constant, and $D_\mathrm{d}$, $D_\mathrm{s}$, and $D_{\mathrm{ds}}$ are the angular diameter distances between the observer and the lens, the observer and the source, and the lens and the source, respectively.

@@

In the special case of an axisymmetric lens we can simplify the notation by considering a single coordinate axis with origin at the centre of the lens. We label the components of the image position, source position and deflection angle along this axis as $\theta$, $\beta$ and $\alpha$, respectively.
The lens equation for an axisymmetric lens then becomes
\begin{equation}\label{eq:1dlenseq}
\beta = \theta - \alpha(\theta),
\end{equation}
and the expression for the deflection angle reduces to
\begin{equation}
\alpha(\theta) = \frac{2}{\theta}\int_0^{\theta} d\theta' \kappa(\theta') \theta'.
\end{equation}
This can also be expressed in terms of the total projected mass enclosed within a circle with angular radius equal to $\theta$:
\begin{equation}
\alpha(\theta) = \frac{1}{\pi\theta}\dfrac{M(<\theta)}{\Sigma_{\mathrm{cr}}D_{\mathrm{d}}^2}.
\end{equation}

A very important quantity for describing the strength of a strong lens is the Einstein radius, $\tein$. For an axisymmetric lens, this is the radius corresponding to the solution of \Eref{eq:1dlenseq} for a source placed at the same angular position as the lens centre ($\beta=0$).
The circle with radius equal to $\tein$ is known as the tangential critical curve. Images that form there have infinite magnification in the tangential direction.
%
It can be shown that $\tein$ satisfies the following condition,
\begin{equation}\label{eq:criteq}
\bar{\kappa}(<\tein) = 1,
\end{equation}
where $\bar{\kappa}(<\theta)$ is the average surface mass density within a radius equal to $\theta$:
\begin{equation}
\bar{\kappa}(<\theta) \equiv \frac{2}{\theta^2}\int_0^{\theta}d\theta' \kappa(\theta')\theta'.
\end{equation}

Axisymmetric lenses of the kind considered in this work 
typically produce either one or three images of a point source.
\Fref{fig:scheme} helps to visualise this property.
Plotted in \Fref{fig:scheme} is the quantity $\theta - \alpha(\theta)$ as a function of the position in the image plane $\theta$, for a few lens models.
According to \Eref{eq:1dlenseq}, images form at the locations where this quantity equals the position of the source. 
Therefore, given a source position $\beta$, the number of images and their location can be determined by drawing a horizontal line at the value $\beta$ on the vertical axis, and finding the points where this line intersects the $\theta - \alpha(\theta)$ curve.

For small values of $\beta$ (for sources close to the lens centre), the number of images that are produced is three: the source is strongly lensed. For large values of $\beta$, instead, only one image is formed.
The value of $\beta$ where the transition occurs is known as the radial caustic, which is marked by the horizontal dotted line in \Fref{fig:scheme}. 
As can be seen from \Fref{fig:scheme}, a source at this location is mapped to the stationary point of the $\theta - \alpha(\theta)$ curve. 
That location on the image plane is known as the radial critical curve. 
Images that form there have a formally infinite magnification in the radial direction\footnote{The slope of the $\theta - \alpha(\theta)$ curve is the inverse of the radial magnification. This can be understood by taking the derivative of \Eref{eq:1dlenseq} with respect to $\theta$.}.

\begin{figure}
\includegraphics[width=\columnwidth]{composite_fixedap_scheme.eps}
\caption{
The lens equation of axisymmetric lenses. Solid lines: right-hand side of \Eref{eq:1dlenseq} for five lenses with different density profiles. 
Given a source at position $\beta$, its lensed images form at the values of $\theta$ where the solid line intersects a horizontal line located at $\beta$ on the vertical axis. 
Dotted lines: positions of the radial caustics. Sources located within the radial caustic of a given lens produce three images.
The lens models used in this simulations consist of a stellar component and a dark matter halo, as described in section \ref{ssec:profile}. Their density profile is plotted in \Fref{fig:kappa}.
\label{fig:scheme}
}
\end{figure}

@@

\Fref{fig:ellcaust} shows the source-plane caustics of elliptical lenses with different values of the axis ratio. We used the software {\sc Glafic} \citep{Ogu21} to obtain the caustic curves.
The outermost curves are radial caustics, while the inner ones are tangential ones.
The most striking difference with respect to the axisymmetric case (blue curves in \Fref{fig:ellcaust}) is the fact that the tangential caustic is transformed from a point into a diamond-like curve.
Sources located within the diamond produce five images (one of which is usually highly de-magnified). Sources that lie in the region enclosed between the two caustics are imaged three times. Sources outside the radial caustic are imaged only once, as in the axisymmetric case.
The fact that the number of images changes by two at a caustic crossing is a general feature of gravitational lenses with no singularities \citep{SEF92}.

@@

All of the examples that we considered were based on sources with a single surface brightness peak.
In general this is not necessarily the case, and the intrinsic number of surface brightness peaks of a strongly lensed source is not known a priori.
When dealing with a real lens candidate, applying our lens definition criterion requires showing that the observed number of surface brightness peaks can be reproduced with a lens model in which the source has a smaller number of peaks.
In samples of lenses defined via visual inspection this process is typically done implicitly, by identifying multiply imaged blobs that belong to the same source element.

To our knowledge, we are the first to propose a surface brightness peak-based definition of a strong lensing event.
A popular alternative definition of a strong lens is one based on magnification: only images that are magnified by more than a given threshold are considered as strongly lensed \citep[see e.g.][]{Hil++07}.
The problem with a definition of this kind is that magnification cannot be determined unambiguously from observations, unless the intrinsic properties of the lensed source are known from other means (e.g. if the source is a standard candle or a standard ruler).
Because of the mass-sheet degeneracy \citep{FGS85}, it is possible to vary the magnification of a lensed image while keeping its observed properties fixed: this could lead to the paradox of two identical-looking lenses that are classified differently on the basis of the underlying magnification.
Although we could still use a magnification-based definition for the sake of carrying out our experiments, it would then be difficult to apply our results to real data.
Our definition of a strong lensing event, instead, is robust with respect to the mass-sheet degeneracy.

@@

In this work we focus on massive early-type galaxies as lenses, as these make up the vast majority of known strong lenses.
We describe their mass distribution with a model consisting of two concentric components, one describing the baryons and one for the dark matter.
We assume that the baryonic component consists entirely of stars, thereby neglecting gas, which is known to contribute very little to the mass of early-type galaxies in the inner regions that are probed by strong lensing. 
We then assume that the stars follow a S\'{e}rsic profile, with 
projected surface mass density given by
\begin{equation}
\Sigma(R) = \Sigma_0 \exp{\left\{-b\left(\frac{R}{\reff}\right)^{1/n}\right\}}.
\end{equation}
In the above equation,
\begin{equation}
\Sigma_0 = \frac{\mstar b^{2n}}{2\pi n \reff^2 \Gamma(2n)},
\end{equation}
$\mstar$ is the total mass, $\reff$ is the radius enclosing half of the total mass, $n$ is the S\'{e}rsic index, $\Gamma$ is the incomplete Gamma function, and $b$ is given by \citep{C+B99}
\begin{equation}
b(n) \approx 2n -\frac13 + \frac{4}{405n} + \frac{46}{25515n^2} + O(n^{-3}).
\end{equation}
Throughout this paper we indicate with $\teff$ the angular size of the half-light radius.

We fix the S\'{e}rsic index of the lenses to $n=4$, corresponding to a de Vaucouleurs model.
Although early-type galaxies are often described with a free S\'{e}rsic index, a de Vaucouleurs profile is able to reproduce their surface brightness distribution to a few percent in the radial range $1\rm{kpc} < R < 30\rm{kpc}$ \citep[see e.g.][]{Son20}, which is the region that is most relevant for strong lensing.
Finally, we assume that the light distribution of the stellar component follows its mass distribution exactly. That is, we do not allow for the presence of gradients in the stellar mass-to-light ratio. In section \ref{ssec:limitations} we discuss qualitatively what the implications of relaxing this assumption would be.

@@

\Fref{fig:kappa} shows the dimensionless surface mass density profile of S\'{e}rsic + gNFW models with various values of the inner dark matter density slope $\gammadm$ and of the fraction of projected dark matter mass within the half-light radius, $\fdm$.
All of the profiles have 
a dark matter scale radius equal to ten times $\reff$, and are normalised in such a way that the Einstein radius is equal to the half-mass radius of the stellar component (these assumptions are dropped later).
Two main features emerge from \Fref{fig:kappa}. First, the baryons generally dominate the total density in the inner regions $(\theta < \tein)$, while the dark matter is the main component at large radii.
Second, models with different dark matter fractions and inner dark matter slopes can conspire to produce very similar total density profiles. 
This is the case, for example, of the ($\fdm=0.5,\gammadm=1.5$) and the ($\fdm=0.2,\gammadm=2.0$) models (green and blue lines in \Fref{fig:kappa}).

@@

Axisymmetric lenses with a density profile of the kind introduced in section \ref{ssec:profile} can produce either one or three images of a point source.
This can be seen in \Fref{fig:scheme}, which shows the lens equation for various values of the dark matter fraction and the inner dark matter slope.
All of the lenses shown in this figure have the same Einstein radius, which is equal in size to the half-mass radius of the stellar component.
The radius of the radial caustic, marked by the dotted lines in \Fref{fig:scheme}, is a strong function of the lens properties: it is largest in lenses with a smaller dark matter fraction or steeper dark matter slopes.
As a result, the source plane area that is subject to strong lensing is an even stronger function of these properties, since it scales with the square of the radial caustic radius.
In order to compute the lensing cross-section, however, we must take into account the magnification, because that determines whether multiple images can be detected or not.

\Fref{fig:1dmag} shows the magnification of the secondary image as a function of source position. The secondary image is the one located in the region between the radial and tangential critical curves, opposite to the source with respect to the lens centre. In most practical cases this is the second brightest image.
As \Fref{fig:1dmag} shows, the magnification is very large for sources close to the lens centre (small values of $\beta$), decreases with increasing source position and then increases in close proximity to the radial caustic.
While for the model with $f_{\mathrm{DM}}=0.8$ (purple curve) the magnification is above unity everywhere, other lens models can produce highly de-magnified secondary images for large values of $\beta$.
Depending on the intrinsic brightness of the lensed source, these images may or may not be detected.
